The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2018-10-01 article.tex
Write v0.7, fix #18

Current phylogenetic tools assume that only a single speciation event can occur at any given time.
While this assumption is useful to construct a wide variety of successful models \richel{@gio: citation here} \gio{@richel: basically all the models we know are based on this assumpion: DDD, PBD, BISSE, MUSSE, SECSSE, any other SSE, DAISIE etc. etc. It's a very very general feature. Maybe being specific could lead the reader to consider things that are, in the end, not essential to the story we want to tell here. Do you still think we need it?},
\richel{Yes, I think here would be a fine spot to cite some of those models, 
I think BiSSE, DAISIE, DDD and PBD would be appropriate}\gio{CLOSED} 
(for example as in \cite{Maddison2007biSSE}, \cite{Valente2015}, \cite{etienne2012diversity}, \cite{etienne2014estimating})
they disallow for environmental changes that trigger speciation on a large scale.

The (constant-rate) birth-death (BD) model embodies the common assumption that 
only a single speciation event can occur at any given time.
The MBD model relaxes this assumption, allowing events in which 
large-scale environmental changes lead to a great number of species 
in relatively short time intervals. Such hypothesis can be useful to describe, for example, 
systems like cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria (\cite{janzen2016}, \cite{janzen2017})\richel{@gio: citation here}\gio{CLOSED}.

\gio{If I described the process in the same way you report in the example I would probably end up writing the same things that we say a few lines below, where we describe the parameters. Don't you think?}
\richel{You described the model in the Methods. I moved it to here}\gio{CLOSED}
In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the usual per-species speciation and extinction rates. 
Additionally, $\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, all species present in the phylogeny at that moment
have a probability $q$ to speciate (independent on $\lambda$).
The number of species that speciate due to this can also be zero. 
\richel{Is this correct?}\gio{this is the code used for the simulation: "stats::rbinom(n = 1, size = N, prob = q)". It can actually yield zero. Now I am wondering if it is correct or not...}.

Unfortunately, a tree prior according to this model, 
providing the probability of a species tree under the MBD model, 
is unavailable in current Bayesian phylogenetic tools. 
Whilst a likelihood equation has been derived (\richel{cite yourself here}\gio{CLOSED})(\cite{mbd}),
it has not been implemented as tree prior yet. 
There are various reasons for this.
First, the computation of the MBD likelihood involves solving a set of 
non-linear differential equations \gio{@richel: are they actually non-linear?}, and while this computation is quite fast, 
it still takes much more time than the corresponding probability 
of the BD model which is a simple analytical formula. 
In a Bayesian MCMC chain, the tree prior probability must be calculated many times, 
and hence the total computation will take considerably longer with a PBD tree prior.

@@

The MBD model has 4 parameters, depicted in table \ref{table:parameters}. 
We pick values of $\nu$ in such a way we expect a multiple speciation
event to be triggered zero ($\nu = 0$), once, twice , four and eight times 
\richel{I assume you can calculate the correct $\nu$} \gio{I dunno if it's possible to do so. This kind of calculation, especially for $\mu>0$ might be very cumbersome. One thing Rampal and I discussed to do was to use an equivalent $\lambda$ in BD to have a similar amount of mutations for both kind of simulated trees. In the code there should be already something like that (it might be not completely accurate though). I explain myself better: a) you simulate with some MBD par setup; b) you count the amount of mutations you had; c) from this number you decide what $lambda$ to use to simulate the BD process to have the (expected) same amount of mutations. It's probably much much easier than going from BD to MBD.}. For each
expected number of triggered events, we only keep those phylogenies
that actually realized the expected number of triggered events.
We pick values of $q$ that are 0.0 (a speciation barrier at the
triggered event), 0.25, 0.5 and 1.0.
We set our extinction rate $\mu$ to 0.1 in all simulation.
As we select our phylogenies on their number of lineages,
we calculate $\lambda$ in a such a way that the mean expected number
of lineages equals the desired numbers of taxa of 50, 100 and 200. 
For $\nu = 0$, the model falls back to a standard BD model.
Note that the $\lambda$ and $q$ have different units
and it is a misconception to think that for $\lambda = q$ (already
impossible due to their units) the MBD model would reduce to a BD model.

We simulate protracted birth-death trees, using the MBD package (\cite{mbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
For each combination of $\lambda, \mu, \nu$ and $q$, 
we generate species trees with a crown age of 15 million years 
\gio{In general [15 million years] is ok for me. Keep in mind, though, that allowing multiple speciations may lead to an explosion in the number of species. Increasing the time by a factor of n usually means increasing the expected number of species at the present by a factor proportional to $e^n$}
\richel{I know}.
Only trees with the desired number of good taxa are kept.

@@

% Bibliography
\gio{bibliography is missing. The only bib file present does not correspond to the bibliography showed in the pdf file.}
\richel{Weird. I see it both locally and on Overleaf. Sent email with my screenshots, show me yours to see if I can help}\gio{CLOSED}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{mee}
\bibliography{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

Current phylogenetic tools assume that only a single speciation event can occur at any given time.
While this assumption is useful to construct a wide variety of successful 
models (for example: \cite{Maddison2007biSSE}, \cite{Valente2015}, \cite{etienne2012diversity}, \cite{etienne2014estimating}),
they disallow for environmental changes that trigger speciations in multiple clades at a same point in time.

The (constant-rate) birth-death (BD) model embodies the common assumption that 
only a single speciation event can occur at any given time.
The MBD model relaxes this assumption, allowing events in which 
large-scale environmental changes lead to a great number of species 
in relatively short time intervals. Such hypothesis can be useful to describe, for example, 
systems like cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria (\cite{janzen2016}, \cite{janzen2017}).

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the usual per-species speciation and extinction rates. 
Additionally, $\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, all species present in the phylogeny at that moment
have a probability $q$ to speciate (independent on $\lambda$).
The number of species that speciate due to this can also be zero. 
\richel{Is this correct?}
\gio{
  this is the code used for the simulation: 
  "stats::rbinom(n = 1, size = N, prob = q)". 
  It can actually yield zero. 
  Now I am wondering if it is correct or not...
}
\richel{
  I read in the PDF describing the likelihood 
  derivation 'The $L^\lambda_m$ term takes into account, being in 
  the state $Q^k_m(t)$, all the possible ways to have at least one birth event from whichever pool'.
  I would suggested changing the code to
  "1 + stats::rbinom(n = 1, size = N, prob = q)"
  and test it somehow to be sure the sim matches the likelihood
}	.

Unfortunately, a tree prior according to this model, 
providing the probability of a species tree under the MBD model, 
is unavailable in current Bayesian phylogenetic tools. 
Whilst a likelihood equation has been derived (\cite{mbd}),
it has not been implemented as tree prior yet. 
There are various reasons for this.
First, the computation of the MBD likelihood involves solving a set of 
non-linear differential equations 
\gio{are they actually non-linear?}
\richel{Definitely, I see an exponentiation and combinatorial term, both are non-linear}, 
and while this computation is quite fast, 
it still takes much more time than the corresponding probability 
of the BD model which is a simple analytical formula. 
In a Bayesian MCMC chain, the tree prior probability must be calculated many times, 
and hence the total computation will take considerably longer with a PBD tree prior.

@@

\richel{This setup is wrong, discuss described setup by Gio today}
The MBD model has 4 parameters, depicted in table \ref{table:parameters}. 
We pick values of $\nu$ in such a way we expect a multiple speciation
event to be triggered zero ($\nu = 0$), once, twice , four and eight times 
\gio{
  One thing Rampal and I discussed to do was to use an 
  equivalent $\lambda$ in BD to have a similar amount of mutations for both kind of simulated trees. 
  In the code there should be already something like 
  that (it might be not completely accurate though). I explain myself 
  better: 
  a) you simulate with some MBD par setup; 
  b) you count the amount of mutations you had; 
  c) from this number you decide what $lambda$ to use to simulate the BD process 
  to have the (expected) same amount of mutations. It's probably much much easier 
  than going from BD to MBD.
}
. For each
expected number of triggered events, we only keep those phylogenies
that actually realized the expected number of triggered events.
We pick values of $q$ that are 0.0 (a speciation barrier at the
triggered event), 0.25, 0.5 and 1.0.
We set our extinction rate $\mu$ to 0.1 in all simulation.
As we select our phylogenies on their number of lineages,
we calculate $\lambda$ in a such a way that the mean expected number
of lineages equals the desired numbers of taxa of 50, 100 and 200. 
For $\nu = 0$, the model falls back to a standard BD model.
Note that the $\lambda$ and $q$ have different units
and it is a misconception to think that for $\lambda = q$ (already
impossible due to their units) the MBD model would reduce to a BD model.

We simulate protracted birth-death trees, using the MBD package (\cite{mbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
For each combination of $\lambda, \mu, \nu$ and $q$, 
we generate species trees with a crown age of 15 million years 
Only trees with the desired number of good taxa are kept.

@@

% Bibliography

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{mee}
\bibliography{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
